---
title: "Chad's 2026 MLB Report"
output: github_document
knit: (function(input, ...) { rmarkdown::render(input, output_file = "README.md", envir = globalenv()) })
---

<!-- *Interested in the underlying code that builds this report?* Check it out on GitHub: [mlb26](https://github.com/chadallison/mlb26){target="_blank"} -->

___

### Contents

Nothing yet!

___

```{r fold_code, include = F}
knitr::knit_hooks$set(source = function(x, options) {
    hook.r = function(x, options) {
      fence = "```"
      language = tolower(options$engine)
      if (language == "node") language = "javascript"
      if (!options$highlight) language = "text"
      if (!is.null(options$fold_code)) {
        paste0("\n\n", "<details><summary>View Code</summary>\n", fence, language,
               "\n", x, fence, "\n\n", "</details>\n")
      } else paste0('\n\n', fence, language, '\n', x, fence,  '\n\n')
    }
    x = knitr:::hilight_source(x, "markdown", options)
    hook.r(paste(c(x, ""), collapse = "\n"), options)
})
```

```{r message = F, warning = F, include = F, fold_code = T}
library(tidyverse)
library(tvthemes)
library(janitor)
library(glue)
library(zoo)
library(baseballr)
library(arrow)

theme_custom = theme_avatar() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, size = 9, vjust = 2.5, face = "italic"),
        plot.caption = element_text(face = "italic"),
        panel.grid.major = element_line(linewidth = 0.5, colour = "#DFDAD1"),
        panel.grid.minor = element_line(linewidth = 0.5, colour = "#DFDAD1"))

theme_set(theme_custom)
```

```{r message = F, warning = F, include = F}
end_games_raw = mlb_schedule(season = 2026, level_ids = "1")

end_games = end_games_raw |>
  filter(series_description == "Regular Season" & status_detailed_state == "Final") |>
  select(date, game_pk, away_score = teams_away_score, away_team = teams_away_team_name,
         home_team = teams_home_team_name, home_score = teams_home_score) |>
  filter(!is.na(away_score) & !is.na(home_score)) |>
  mutate(win_team = ifelse(home_score > away_score, home_team, away_team),
         lose_team = ifelse(home_score > away_score, away_team, home_team),
         win_score = ifelse(home_score > away_score, home_score, away_score),
         lose_score = ifelse(home_score > away_score, away_score, home_score))

# writing and reading prevents baseballr message
write_csv(end_games, "data/end_games.csv")
end_games = read_csv("data/end_games.csv", show_col_types = F)
```

```{r include = F}
all_teams = sort(unique(c(end_games$home_team, end_games$away_team)))

team_abbrevs = c(
  "ARI", "ATH", "ATL", "BAL", "BOS", "CHC",
  "CWS", "CIN", "CLE", "COL", "DET", "HOU",
  "KC", "LAA", "LAD", "MIA", "MIL", "MIN",
  "NYM", "NYY", "PHI", "PIT", "SD", "SF",
  "SEA", "STL", "TB", "TEX", "TOR", "WSH"
)

team_hex = c(
  "#A71930", "#003831", "#CE1141", "#DF4601", "#BD3039", "#0E3386",
  "#27251F", "#C6011F", "#0C2340", "#33006F", "#FA4616", "#EB6E1F",
  "#004687", "#BA0021", "#005A9C", "#00A3E0", "#FFC52F", "#D31145",
  "#002D72", "#003087", "#E81828", "#FDB827", "#2F241D", "#FD5A1E",
  "#0C2C56", "#C41E3A", "#8FBCE6", "#C0111F", "#134A8E", "#dcbdd0"
)

divisions = c(
  "NL West", "AL West", "NL East", "AL East", "AL East", "NL Central",
  "AL Central", "NL Central", "AL Central", "NL West", "AL Central", "AL West",
  "AL Central", "AL West", "NL West", "NL East", "NL Central", "AL Central",
  "NL East", "AL East", "NL East", "NL Central", "NL West", "NL West",
  "AL West", "NL Central", "AL East", "AL West", "AL East", "NL East"
)

team_divisons = data.frame(team = all_teams, division = divisions) |>
  mutate(division = factor(division, levels = c("AL West", "AL Central", "AL East",
                                                "NL West", "NL Central", "NL East")))

teams_info = data.frame(team = all_teams, abb = team_abbrevs, hex = team_hex)

better_date = function(dt) {
  return(as.character(glue("{month(dt, label = T, abbr = T)} {day(dt)}, {year(dt)}")))
}

today_nice = better_date(Sys.Date())
```

```{r}
all_results = end_games |>
  transmute(
    date, game_pk,
    team = home_team, opp = away_team,
    score = home_score, opp_score = away_score,
    is_win = ifelse(home_score > away_score, 1, 0)
  ) |>
  bind_rows(
    end_games |>
      transmute(
        date, game_pk,
        team = away_team, opp = home_team,
        score = away_score, opp_score = home_score,
        is_win = ifelse(away_score > home_score, 1, 0)
      )
  ) |>
  arrange(team, date)

team_records = all_results |>
  group_by(team) |>
  summarise(gp = n(),
            scored = sum(score),
            allowed = sum(opp_score),
            run_diff = sum(score) - sum(opp_score),
            wins = sum(is_win),
            losses = sum(1 - is_win),
            win_pct = mean(is_win)) |>
  mutate(record = paste0(wins, "-", losses)) |>
  inner_join(teams_info, by = "team")
```

```{r}
team_records |>
  arrange(desc(win_pct), desc(run_diff), desc(scored), allowed, team) |>
  mutate(pct_vs_500 = win_pct - 0.5,
         pl = ifelse(pct_vs_500 >= 0, record, ""),
         nl = ifelse(pct_vs_500 < 0, record, ""),
         row_num = row_number()) |>
  ggplot(aes(reorder(team, -row_num), pct_vs_500)) +
  geom_col(aes(fill = hex)) +
  geom_text(aes(label = pl), hjust = -0.25, size = 3) +
  geom_text(aes(label = nl), hjust = 1.25, size = 3) +
  coord_flip() +
  scale_fill_identity() +
  labs(title = glue("Team Standings as of {today_nice}"),
       x = NULL, y = "Team win pct. over/under .500") +
  scale_y_continuous(
    breaks = seq(-0.5, 0.5, by = 0.1), labels = scales::percent,
    expand = expansion(mult = 0.1)
  )
```

___

```{r}
plot_data = team_records |>
  mutate(run_diff_pg = round(run_diff / gp, 2)) |>
  arrange(desc(run_diff_pg), desc(scored), allowed, team) |>
  mutate(row_num = row_number(),
         pl = ifelse(run_diff_pg >= 0, run_diff_pg, ""),
         nl = ifelse(run_diff_pg < 0, run_diff_pg, ""))

plot_data |>
  ggplot(aes(reorder(team, -row_num), run_diff_pg)) +
  geom_col(aes(fill = hex)) +
  geom_text(aes(label = pl), hjust = -0.25, size = 3) +
  geom_text(aes(label = nl), hjust = 1.25, size = 3) +
  scale_fill_identity() +
  scale_y_continuous(breaks = seq(-10, 10, by = 0.5)) +
  coord_flip(ylim = c(min(plot_data$run_diff_pg) * 1.05, max(plot_data$run_diff_pg) * 1.05)) +
  labs(title = glue("Avg. Run Differentials as of {today_nice}"),
       x = NULL, y = "Avg. Run differential")
```

```{r}
team_rpg = all_results |>
  group_by(team) |>
  summarise(gp = n(),
            off_rpg = mean(score),
            def_rpg = mean(opp_score))

avg_runs = mean(all_results$score)

team_rpg |>
  inner_join(teams_info, by = "team") |>
  ggplot(aes(off_rpg, def_rpg)) +
  geom_point(aes(col = hex), shape = "square", size = 4) +
  ggrepel::geom_text_repel(aes(label = abb), size = 3, max.overlaps = 30) +
  scale_color_identity() +
  geom_vline(xintercept = avg_runs, linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = avg_runs, linetype = "dashed", alpha = 0.5) +
  scale_x_continuous(breaks = seq(0, 100, by = 0.5)) +
  scale_y_continuous(breaks = seq(0, 100, by = 0.5)) +
  labs(x = "Runs scored per game",
       y = "Runs allowed per game",
       title = glue("Runs scored/allowed per game by team as of {today_nice}"))
```

```{r}
team_pythag = all_results |>
  group_by(team) |>
  summarise(scored = sum(score),
            allowed = sum(opp_score)) |>
  mutate(py = scored ^ 2 / (scored ^ 2 + allowed ^ 2),
         py_500 = py - 0.5)

team_pythag |>
  mutate_if(is.numeric, round, 3) |>
  mutate(py_lab = paste0(py * 100, "%")) |>
  inner_join(teams_info, by = "team") |>
  mutate(pl = ifelse(py_500 >= 0, py_lab, ""),
         nl = ifelse(py_500 < 0, py_lab, "")) |>
  ggplot(aes(reorder(team, py), py_500)) +
  geom_col(aes(fill = hex)) +
  geom_text(aes(label = pl), size = 3, hjust = -0.25) +
  geom_text(aes(label = nl), size = 3, hjust = 1.25) +
  coord_flip(ylim = c(min(team_pythag$py_500) * 1.1, max(team_pythag$py_500) * 1.1)) +
  scale_fill_identity() +
  labs(x = NULL, y = "Pythagorean win percentage vs. .500",
       title = "Team Pythagorean Wins") +
  scale_y_continuous(breaks = seq(-1, 1, by = 0.1), labels = scales::percent)
```

```{r}
plot_data = all_results |>
  mutate(raw_diff = score - opp_score,
         diff_sign = sign(raw_diff),
         abs_diff = abs(raw_diff),
         adj_abs_diff = sqrt(abs_diff),
         adj_diff = adj_abs_diff * diff_sign) |>
  group_by(team) |>
  summarise(gp = n(),
            total_adj_diff = sum(adj_diff),
            avg_adj_diff = mean(adj_diff)) |>
  mutate(scaled_diff = scale(avg_adj_diff),
         lab = round(scaled_diff, 2),
         pl = ifelse(scaled_diff >= 0, lab, ""),
         nl = ifelse(scaled_diff < 0, lab, "")) |>
  inner_join(teams_info, by = "team")

plot_data |>
  ggplot(aes(reorder(team, scaled_diff), scaled_diff)) +
  geom_col(aes(fill = hex)) +
  scale_fill_identity() +
  coord_flip(ylim = c(min(plot_data$scaled_diff) * 1.05, max(plot_data$scaled_diff) * 1.05)) +
  geom_text(aes(label = pl), size = 3, hjust = -0.25) +
  geom_text(aes(label = nl), size = 3, hjust = 1.25) +
  labs(x = NULL, y = "Adjusted run differential",
       title = glue("Adjusted Run Differentials as of {today_nice}")) +
  scale_y_continuous(breaks = seq(-3, 3, by = 0.5))
```

```{r}
end_with_npr = end_games |>
  inner_join(team_rpg, by = c("home_team" = "team")) |>
  rename(home_gp = gp, home_off_rpg = off_rpg, home_def_rpg = def_rpg) |>
  inner_join(team_rpg, by = c("away_team" = "team")) |>
  rename(away_gp = gp, away_off_rpg = off_rpg, away_def_rpg = def_rpg) |>
  mutate(home_exp = (home_off_rpg + away_def_rpg) / 2,
         away_exp = (away_off_rpg + home_def_rpg) / 2,
         home_off_npr = home_score - home_exp,
         home_def_npr = away_exp - away_score,
         away_off_npr = away_score - away_exp,
         away_def_npr = home_exp - home_score)

plot_data = end_with_npr |>
  select(date, team = home_team, opp = away_team, off_npr = home_off_npr, def_npr = home_def_npr) |>
  bind_rows(
    end_with_npr |>
      select(date, team = away_team, opp = home_team, off_npr = away_off_npr, def_npr = away_def_npr)
  ) |>
  arrange(team, date) |>
  group_by(team) |>
  summarise(gp = n(),
            net_off_npr = sum(off_npr),
            net_def_npr = sum(def_npr),
            net_ovr_npr = sum(off_npr) + sum(def_npr),
            avg_off_npr = mean(off_npr),
            avg_def_npr = mean(def_npr),
            avg_ovr_npr = mean(off_npr) + mean(def_npr)) |>
  mutate(avg_npr_scaled = scale(avg_ovr_npr),
         lab = round(avg_npr_scaled, 3),
         pl = ifelse(avg_npr_scaled >= 0, lab, ""),
         nl = ifelse(avg_npr_scaled < 0, lab, "")) |>
  inner_join(teams_info, by = "team")

plot_data |>
  ggplot(aes(reorder(team, avg_ovr_npr), avg_ovr_npr)) +
  geom_col(aes(fill = hex)) +
  scale_fill_identity() +
  coord_flip(ylim = c(min(plot_data$avg_ovr_npr) * 1.05, max(plot_data$avg_ovr_npr) * 1.05)) +
  geom_text(aes(label = pl), size = 3, hjust = -0.25) +
  geom_text(aes(label = nl), size = 3, hjust = 1.25) +
  labs(x = NULL, y = "Avg. NPR",
       title = glue("Team Avg. NPR as of {today_nice}")) +
  scale_y_continuous(breaks = seq(-5, 5, by = 0.5))
```



```{r}
all_results |>
  mutate(game_num = row_number(),
         gp = max(game_num),
         .by = "team") |>
  # filter(game_num > gp - 10) |>
  mutate(sc = sqrt(score),
         al = sqrt(opp_score)) |>
  group_by(team) |>
  summarise(sc = sum(sc),
            al = sum(al)) |>
  mutate(py = sc ^ 2 / (sc ^ 2 + al ^ 2),
         wins = round(py * 162, 1)) |>
  arrange(desc(py))
```


```{r}
all_results |>
  mutate(diff = score - opp_score,
         scaled = scale(diff)) |>
  group_by(team) |>
  summarise(score = sum(scaled),
            avg_score = mean(scaled)) |>
  arrange(desc(avg_score)) |>
  mutate_if(is.numeric, round, 3) |>
  mutate(pl = ifelse(avg_score >= 0, avg_score, ""),
         nl = ifelse(avg_score < 0, avg_score, "")) |>
  inner_join(teams_info, by = "team") |>
  ggplot(aes(reorder(team, avg_score), avg_score)) +
  geom_col(aes(fill = hex)) +
  scale_fill_identity() +
  coord_flip() +
  scale_y_continuous(breaks = seq(-3, 3, by = 0.1), expand = expansion(mult = 0.1)) +
  geom_text(aes(label = pl), size = 3, hjust = -0.25) +
  geom_text(aes(label = nl), size = 3, hjust = 1.25) +
  labs(x = NULL, y = "Avg. Rating",
       title = glue("Scaled Team Ratings as of {today_nice}"))
```
























































